Dynamics of Colombo's Top: Tidal Dissipation and Resonance Capture

### 2021-01-10 writeups/Su_weak_tides.tex
rm bad footnote per ref request
stash beginning work on weak tides

    \newcommand*{\rd}[2]{\frac{\mathrm{d}#1}{\mathrm{d}#2}}
    \newcommand*{\rtd}[2]{\frac{\mathrm{d}^2#1}{\mathrm{d}#2^2}}
    \newcommand*{\pd}[2]{\frac{\partial#1}{\partial#2}}
    \newcommand*{\rdil}[2]{\mathrm{d}#1 / \mathrm{d}#2}
    \newcommand*{\pdil}[2]{\partial#1 / \partial#2}
    \newcommand*{\md}[2]{\frac{\mathrm{D}#1}{\mathrm{D}#2}}
    \newcommand*{\at}[1]{\left.#1\right|}
    \newcommand*{\abs}[1]{\left|#1\right|}
    \newcommand*{\ev}[1]{\langle#1\rangle}
    \newcommand*{\bsmb}[1]{\boldsymbol{\mathbf{#1}}}
    \newcommand*{\uv}[1]{\hat{\bsmb{#1}}}
    \newcommand*{\p}[1]{\left(#1\right)}
    \newcommand*{\s}[1]{\left[#1\right]}
    \newcommand*{\z}[1]{\left\{#1\right\}}
    \DeclareMathOperator*{\argmin}{argmin}
    \DeclareMathOperator*{\argmax}{argmax}
    \DeclareMathOperator*{\med}{med}

@@

Separatrix crossing was studied by \citep{henrard1982}.

@@

In this section, we first briefly lay out the spin dynamics of the planet,
introducing the Cassini State spin-orbit resonance \citep[for more details,
see][]{su2020}. We then introduce the equilibrium tide model of weak tidal
friction used in this work, though much of our analysis does not depend on the
detailed tidal model used.

@@

We study an oblate planet orbiting around a star hosting a further out planet.
The equations of motion in the absence of dissipation are:
\begin{align}
    \rd{\hat{s}}{t}
        &= \omega_{s1}\p{\hat{s} \cdot \hat{l}_1}\p{\hat{s} \times \hat{l}_1}
            - \omega_{1p}\cos I\p{\hat{s} \times \hat{l}_p},\\
    \omega_{s1} &= \frac{3k_q}{2k}\frac{M_*}{m_1}\p{\frac{R_1}{a_1}}^3 s,\\
    \omega_{1p} &= \frac{3m_p}{4m_*}\p{\frac{a_1}{a_p\sqrt{1 - e_p^2}}}^3 \Omega_1.
\end{align}

We define $s_c$ to be the critical spin where the precession frequencies
$\omega_{s1}$ and $\omega_{1p}$ are equal, or
\begin{equation}
    \at{\omega_{s1}}_{s = s_c} = \omega_{1p}\cos I.
\end{equation}
rhroughout this paper, we often set the orbital frequency of the inner planet
$\Omega_1 = 1$.

@@

Equilibria of spin dynamics are \emph{Cassini States} (CSs). Refer to Su \& Lai
2020 for more detailed discussion. In the notation of the previous paper,
\begin{equation}
    \eta \equiv -\frac{g}{\alpha} = s_c / s.
\end{equation}

Plots of CS locations as always in \autoref{fig:cs_locs}.
\begin{figure}
    \centering
    \includegraphics[width=\columnwidth]{../initial/99_misc/2_cs_locs.png}
    \caption{CS locations, TODO change axis labels etc.}\label{fig:cs_locs}
\end{figure}

Note that $\phi$ is conjugate to $\cos \theta$, and thus the spin dynamics
exhibit Hamiltonian:
\begin{equation}
    H(\mu, \phi; s) = -\frac{s}{s_c}\frac{\mu^2}{2}
        + \mu \cos I - \sin I \sqrt{1 - \mu^2}\cos \phi.
\end{equation}

Plot of level curves of Hamiltonian as before, including $\mathcal{C}_{\pm}$
notation in \autoref{fig:1contours}.
\begin{figure}
    \centering
    \includegraphics[width=\columnwidth]{../initial/0_eta/1contours.png}
    \caption{TODO label $\mathcal{C}_{\pm}$.}\label{fig:1contours}
\end{figure}

@@

We use the weak tidal friction model \citep{lai2012}. The EOM for $\theta, s$
become:
\begin{align}
    \rd{\hat{s}}{\tau}
        &= \frac{s}{s_c}\p{\hat{s} \cdot \hat{l}_1}\p{\hat{s} \times
                \hat{l}_1} - \hat{s} \times \hat{l}_p
            + \frac{\epsilon 2\Omega_1}{s}
                \p{1 - \frac{s}{2\Omega_1}\p{\hat{l}_1 \cdot \hat{s}}}
                    \hat{s} \times \p{\hat{l}_1 \times \hat{s}},\\
    \rd{s}{\tau}
        &= \epsilon 2\Omega_1 \p{\hat{s} \cdot \hat{l}_1 -
            \frac{s}{2\Omega_1}\p{1 + \p{\hat{s} \cdot \hat{l}_1}^2}}.
\end{align}

@@

As a result of \autoref{ss:tidal_eqs}, the tCE are the only possible final
outcomes. We generate initial spin vectors $\hat{s}$ for isotropic initial
distribution $\p{\cos \theta, \phi}$, and average over $\phi$ in histograms.
This gives histograms in
Figs.~\ref{fig:Hhists_0_06,fig:Hhists_0_20,fig:Hhists_0_70}
\begin{figure}
    \centering
    \includegraphics[width=\columnwidth]{../initial/1_weaktide/5Hhists0_06_5.png}
    \caption{$s_c = 0.06$. Note that it is difficult to reach
    tCE2.}\label{fig:Hhists_0_06}
\end{figure}
\begin{figure}
    \centering
    \includegraphics[width=\columnwidth]{../initial/1_weaktide/5Hhists0_20_5.png}
    \caption{$s_c = 0.2$. Note that tCE2 is both reached with substantial
    probability and has substantial obliquity.}\label{fig:Hhists_0_20}
\end{figure}
\begin{figure}
    \centering
    \includegraphics[width=\columnwidth]{../initial/1_weaktide/5Hhists0_60_5.png}
    \caption{$s_c = 0.6$. Note that tCE2 becomes attracting over tCE1, but will
    have obliquity $\approx I$ and is uninteresting.}\label{fig:Hhists_0_70}
\end{figure}

Similarly to \citet{su2020}, these probabilities are the result of separatrix
crossings. However, the calculation differs from \citet{su2020}: since the
perturbation is dissipative in phase space (i.e.\ it does not conserve phase
space area), analysis following \citet{henrard1982} is not sufficient. A more
general theory of separatrix crossing indeed lets us semi-analytically calculate
the outcome probabilities. Appendix~\ref{app:probs} gives a more thorough
overview of separatrix crossing dynamics, but the key result is:
\begin{equation}
    a^2 + b^2 = c^2.
\end{equation}
We can apply this curve to calculate the likelihood of the tCEs when a
particular trajectory encounters the separatrix, and from this obtain a
semi-analytical prediction of the distribution of outcomes. The semi-analytical
nature of this calculation arises because $\eta_\star$ cannot be calculated
\emph{a priori}. The agreement of this curve with the histograms is examined for
the same three $s_c$ values in
Figs.~\ref{fig:pc_fits_0_06,fig:pc_fits_0_20,fig:pc_fits_0_70}.
\begin{figure}
    \centering
    \includegraphics[width=\columnwidth]{../initial/1_weaktide/5pc_fits0_06_5.png}
    \caption{$s_c = 0.06$ prediction vs histogram.}\label{fig:pc_fits_0_06}
\end{figure}
\begin{figure}
    \centering
    \includegraphics[width=\columnwidth]{../initial/1_weaktide/5pc_fits0_20_5.png}
    \caption{$s_c = 0.20$ prediction vs histogram.}\label{fig:pc_fits_0_20}
\end{figure}
\begin{figure}
    \centering
    \includegraphics[width=\columnwidth]{../initial/1_weaktide/5pc_fits0_70_5.png}
    \caption{$s_c = 0.70$ prediction vs histogram.}\label{fig:pc_fits_0_70}
\end{figure}

@@

Note that while only the results for an isotropic distribution of initial
$\hat{s}$ are shown, in principle arbitrary distributions $P\p{\theta_i}$ can be
convolved against the $P_{tCE2}\p{\theta_i}$ distributions shown in
\autoref{ss:q_dist}.

>>>>

    \newcommand*{\uv}[1]{\hat{\bsmb{#1}}}
    \newcommand*{\rd}[2]{\frac{\mathrm{d}#1}{\mathrm{d}#2}}
    \newcommand*{\rtd}[2]{\frac{\mathrm{d}^2#1}{\mathrm{d}#2^2}}
    \newcommand*{\pd}[2]{\frac{\partial#1}{\partial#2}}
    \newcommand*{\rdil}[2]{\mathrm{d}#1 / \mathrm{d}#2}
    \newcommand*{\pdil}[2]{\partial#1 / \partial#2}
    \newcommand*{\md}[2]{\frac{\mathrm{D}#1}{\mathrm{D}#2}}
    \newcommand*{\at}[1]{\left.#1\right|}
    \newcommand*{\abs}[1]{\left|#1\right|}
    \newcommand*{\ev}[1]{\langle#1\rangle}
    \newcommand*{\bsmb}[1]{\boldsymbol{\mathbf{#1}}}
    \newcommand*{\p}[1]{\left(#1\right)}
    \newcommand*{\s}[1]{\left[#1\right]}
    \newcommand*{\z}[1]{\left\{#1\right\}}
    \DeclareMathOperator*{\argmin}{argmin}
    \DeclareMathOperator*{\argmax}{argmax}
    \DeclareMathOperator*{\med}{med}

@@

\begin{itemize}
    \item Studying planetary obliquities is important. Cassini States are key.

    \item In CS, resonance when two frequencies $g$ and $\alpha \cos \theta$
        become commensurate, $g \simeq \alpha \cos \theta$, where $\theta$ is
        the planetary obliquity.

    \item Many works consider entry into a CS by having $g$ or $\alpha$ vary
        over time to effect resonance capture \citep[e.g.][]{anderson2018teeter,
        millholland_disk, su2020, anderson2020excitation}. However, a third
        possibility exists: $\theta$ can evolve and sweep through a resonance,
        e.g.\ under tidal friction.

    \item Resonance capture via separatrix crossing was first considered by
        \citep{henrard1982} for non-dissipative perturbations
        \citep[e.g.][]{su2020}. However, tidal friction is dissipative, so this
        formalism does not apply. We show that this generalizes.
\end{itemize}

@@

In this section, we first briefly lay out the spin dynamics of the planet,
introducing the Cassini State spin-orbit resonance \citep[for more details,
see][]{su2020}. We then introduce the equilibrium tide model of weak tidal
friction used in this work \citep{lai2012}. While there are many models of tidal
dissipation that may be more accurate (e.g. CITE), our qualitative conclusions
do not depend on the specific model used, so we use the equilibrium tide model
for simplicity.

@@

We consider a star of mass $M_\star$ hosting an inner oblate planet of mass $m$
and radius $R$ with semi-major axis $a$ and an outer planet of mass $m_{\rm p}$
with semi-major axis $a_{\rm p}$ and eccentricity $e_{\rm p}$. Let the inner
planet be spinning with spin $\bsmb{s}$ (with $s \equiv \abs{\bsmb{s}}$ the spin
angular velocity and $\uv{s}$ the spin orientation), and call the unit angular
momenta of the inner and outer planets' orbits $\uv{l}$ and $\uv{l}_{\rm p}$
respectively. The equations of motion in the absence of tidal friction are:
\begin{align}
    \rd{\uv{s}}{t}
        &= \omega_{\rm sl}\p{\uv{s} \cdot \uv{l}}\p{\uv{s} \times \uv{l}}
            - \omega_{\rm lp}\cos I\p{\uv{s} \times \uv{l}_{\rm p}},\\
    \omega_{\rm sl} &\equiv \frac{3GJ_2 mR^2 M_\star}{2a^3 I\Omega}
        = \frac{3k_q}{2k}\frac{M_\star}{m}\p{\frac{R}{a}}^3 s,
            \label{eq:wsl}\\
    \omega_{\rm lp} &= \frac{3m_p}{4M_\star}\p{\frac{a}{a_{\rm p}\sqrt{1 -
            e_{\rm p}^2}}}^3 n.\label{eq:wlp}
\end{align}
In Eq.~\eqref{eq:wsl}, $I = k mR^2$ (with $k$ a constant) is the moment of
inertia and $J_2=k_{\rm q}s^2 (R^3/Gm)$ (with $k_{q}$ a constant) the
rotation-induced (dimensionless) quadrupole of the planet [for a body with
uniform density, $k=0.4, k_{q}=0.5$; for rocky planets, $k\simeq 0.2$ and
$k_{qp}\simeq 0.17$ \citep[e.g.][]{lainey2016quantification} \textcolor{Corr}{?
not sure}]. In other studies, $3k_{\rm qp} / 2 k_{\rm p}$ is often notated as
$k_2 / 2C$ \citep[e.g.][]{millholland_disk}. In Eq.~\eqref{eq:wlp}, $n \equiv
\sqrt{GM_\star/a^3}$ is the inner planet's orbital mean motion, and we have
assumed $a_{\rm p}\gg a$ and included only the leading-order (quadrupole)
interaction between the planet and disk. Following standard notation, we define
$\alpha = \omega_{\rm sl}$ and $g \equiv -\omega_{\rm 1p} \cos I$
\citep[e.g.][]{colombo1966}.

We define $s_{\rm c}$ to be the critical spin where
\begin{equation}
    \at{\alpha}_{s = s_{\rm c}} = -g.
\end{equation}

@@

Equilibria of spin dynamics are \emph{Cassini States} (CSs). Refer to Su \& Lai
2020 for more detailed discussion. In the notation of the previous paper,
\begin{equation}
    \eta \equiv -\frac{g}{\alpha} = \frac{s_{\rm c}}{s}.
\end{equation}

Plots of CS locations as always in Fig.~\ref{fig:cs_locs}.
\begin{figure}
    \centering
    \includegraphics[width=\columnwidth]{../initial/99_misc/2_cs_locs.png}
    \caption{CS locations, TODO change axis labels etc.}\label{fig:cs_locs}
\end{figure}

Note that $\phi$ is conjugate to $\cos \theta$, and thus the spin dynamics
obey Hamiltonian:
\begin{equation}
    H(\mu, \phi; s) = -\frac{s}{s_{\rm c}}\frac{\mu^2}{2}
        + \mu \cos I - \sin I \sqrt{1 - \mu^2}\cos \phi.
\end{equation}

Plot of level curves of Hamiltonian as before, including $\mathcal{C}_{\pm}$
notation in Fig.~\ref{fig:1contours}.
\begin{figure}
    \centering
    \includegraphics[width=\columnwidth]{../initial/0_eta/1contours.png}
    \caption{TODO label $\mathcal{C}_{\pm}$.}\label{fig:1contours}
\end{figure}

@@

We implement the weak friction theory of equilibrium tides \citep{lai2012},
under which both $\theta$ and $s$ evolve. We define the dimensionless time
\begin{equation}
    \tau \equiv gt.
\end{equation}
The full equations of motion including weak tidal friction are then
\begin{align}
    \rd{\uv{s}}{\tau}
        ={}& \frac{s}{s_c}\p{\uv{s} \cdot \uv{l}}\p{\uv{s} \times
                \uv{l}} - \uv{s} \times \uv{l}_{\rm p}\nonumber\\
            &+ \frac{\epsilon 2n}{s}
                \p{1 - \frac{s}{2n}\p{\uv{l} \cdot \uv{s}}}
                    \uv{s} \times \p{\uv{l} \times \uv{s}},\\
    \rd{s}{\tau}
        ={}& \epsilon 2n \p{\uv{s} \cdot \uv{l} -
            \frac{s}{2n}\p{1 + \p{\uv{s} \cdot \uv{l}}^2}},
\end{align}
where $\epsilon$ measures the tidal dissipation rate and is given by
\begin{align}
    \epsilon &\equiv -\frac{1}{gt_a}\frac{L}{2S} \frac{s}{2n},\\
    \frac{1}{t_a} &= \frac{3k_2}{Q}\p{\frac{m}{M}}\p{\frac{R}{a}}^5 n.
\end{align}
Here, $L$ and $S$ are the orbital and spin angular momenta of the inner planet,
respectively.

@@

As a result of Section~\ref{ss:tidal_eqs}, the tCE are the only possible final
outcomes. We generate initial spin vectors $\hat{s}$ for isotropic initial
distribution $\p{\cos \theta, \phi}$, and average over $\phi$ in histograms.
This gives histograms in
Figs.~\ref{fig:Hhists_0_06,fig:Hhists_0_20,fig:Hhists_0_70}
\begin{figure}
    \centering
    \includegraphics[width=\columnwidth]{../initial/1_weaktide/5Hhists0_06_5.png}
    \caption{$s_c = 0.06$. Note that it is difficult to reach
    tCE2.}\label{fig:Hhists_0_06}
\end{figure}
\begin{figure}
    \centering
    \includegraphics[width=\columnwidth]{../initial/1_weaktide/5Hhists0_20_5.png}
    \caption{$s_c = 0.2$. Note that tCE2 is both reached with substantial
    probability and has substantial obliquity.}\label{fig:Hhists_0_20}
\end{figure}
\begin{figure}
    \centering
    \includegraphics[width=\columnwidth]{../initial/1_weaktide/5Hhists0_60_5.png}
    \caption{$s_c = 0.6$. Note that tCE2 becomes attracting over tCE1, but will
    have obliquity $\approx I$ and is uninteresting.}\label{fig:Hhists_0_70}
\end{figure}

Similarly to \citet{su2020}, these probabilities are the result of separatrix
crossings. However, the calculation differs from \citet{su2020}: since the
perturbation is dissipative in phase space (i.e.\ it does not conserve phase
space area), analysis following \citet{henrard1982} is not sufficient. A more
general theory of separatrix crossing is still able to predict the outcome
probabilities. Appendix~\ref{app:probs} gives a more thorough overview of
separatrix crossing dynamics, but the key result is:
\begin{equation}
    a^2 + b^2 = c^2.
\end{equation}
We can apply this curve to calculate the likelihood of the tCEs when a
particular trajectory encounters the separatrix, and from this obtain a
semi-analytical prediction of the distribution of outcomes. The semi-analytical
nature of this calculation arises because $\eta_\star$ cannot be calculated
\emph{a priori}. The agreement of this curve with the histograms is examined for
the same three $s_c$ values in
Figs.~\ref{fig:pc_fits_0_06,fig:pc_fits_0_20,fig:pc_fits_0_70}.
\begin{figure}
    \centering
    \includegraphics[width=\columnwidth]{../initial/1_weaktide/5pc_fits0_06_5.png}
    \caption{$s_c = 0.06$ prediction vs histogram.}\label{fig:pc_fits_0_06}
\end{figure}
\begin{figure}
    \centering
    \includegraphics[width=\columnwidth]{../initial/1_weaktide/5pc_fits0_20_5.png}
    \caption{$s_c = 0.20$ prediction vs histogram.}\label{fig:pc_fits_0_20}
\end{figure}
\begin{figure}
    \centering
    \includegraphics[width=\columnwidth]{../initial/1_weaktide/5pc_fits0_70_5.png}
    \caption{$s_c = 0.70$ prediction vs histogram.}\label{fig:pc_fits_0_70}
\end{figure}

@@

Note that while only the results for an isotropic distribution of initial
$\hat{s}$ are shown, in principle arbitrary distributions $P\p{\theta_i}$ can be
convolved against the $P_{tCE2}\p{\theta_i}$ distributions shown in
Section~\ref{ss:q_dist}.
